---
title: "Reproduction of Chakraborty 2021: An intracategorical analysis of COVID-19 and people with disabilities"
author: "Dermot McMillan"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: sentence
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../docs/report") })
---

# Abstract

This study is a Replication of research done by Chakraborty (2021). Chakraborty analyses county level data to find trends between disability rates and COVID-19 infection rates. The study uses Bivariate Pearson product moment correlation and Generalized Estimating functions to test for the aforementioned correlations in 18 different socio-demographic categories. The purpose of the study is to analyze whether people with disabilities face disproportionate outcomes from COVID-19. T

This study is a *replication* of:

Chakraborty, J.
2021.
Social inequities in the distribution of COVID-19: An intra-categorical analysis of people with disabilities in the U.S.
*Disability and Health Journal* 14:1-5.
<https://doi.org/10.1016/j.dhjo.2020.101007>

# Study metadata

- `Key words`: COVID-19, Disability, Intersectionality, Race/ethnicity, Poverty, Reproducibility
- `Subject`: select from the [BePress Taxonomy](http://digitalcommons.bepress.com/cgi/viewcontent.cgi?article=1008&context=reference)
- `Date created`: date when project was started
- `Date modified`: date of most recent revision
- `Spatial Coverage`: Specify the geographic extent of your study. This may be a place name and link to a feature in a gazetteer like GeoNames or OpenStreetMap, or a well known text (WKT) representation of a bounding box.
- `Spatial Resolution`: Specify the spatial resolution as a scale factor, description of the level of detail of each unit of observation (including administrative level of administrative areas), and/or or distance of a raster GRID size
- `Spatial Reference System`: Specify the geographic or projected coordinate system for the study, e.g. EPSG:4326
- `Temporal Coverage`: Specify the temporal extent of your study---i.e. the range of time represented by the data observations.
- `Temporal Resolution`: Specify the temporal resolution of your study---i.e. the duration of time for which each observation represents or the revisit period for repeated observations
- `Funding Name`: name of funding for the project
- `Funding Title`: title of project grant
- `Award info URI`: web address for award information
- `Award number`: award number

## Original study spatio-temporal metadata

- `Spatial Coverage`: extent of original study
- `Spatial Resolution`: resolution of original study
- `Spatial Reference System`: spatial reference system of original study
- `Temporal Coverage`: temporal extent of original study
- `Temporal Resolution`: temporal resolution of original study

# Study design

The aim of this reproduction study is to implement the original study as closely as possible to reproduce the map of county level distribution of COVID-19 incidence rate, the summary statistics and bivariate correlation for disability characteristics and COVID-19 incidence.

# Materials and procedure

## Computational environment

```{r environment-setup, include = FALSE}
# record all the packages you are using here
# this includes any calls to library(), require(),
# and double colons such as here::i_am()
packages <- c(
  "tidycensus", "tidyverse", "downloader", "sf", "classInt", "readr",
  "here", "s2", "pastecs", "tmap", "SpatialEpi", "svDialogs",
  "geepack", "knitr", "kableExtra", "foreign", "broom", "dotwhisker", "dotenv"
)
# force all conflicts to become errors
# if you load dplyr and use filter(), R has to guess whether you mean dplyr::filter() or stats::filter()
# the conflicted package forces you to be explicit about this
# disable at your own peril
# https://conflicted.r-lib.org/
require(conflicted)

# load and install required packages
# https://groundhogr.com/
if (!require(groundhog)) {
  install.packages("groundhog")
  require(groundhog)
}

# this date will be used to determine the versions of R and your packages
# it is best practice to keep R and its packages up to date
groundhog.day <- "2025-02-25"

# this replaces any library() or require() calls
groundhog.library(packages, groundhog.day)
# you may need to install a correct version of R
# you may need to respond OK in the console to permit groundhog to install packages
# you may need to restart R and rerun this code to load installed packages
# In RStudio, restart r with Session -> Restart Session

# record the R processing environment
# alternatively, use devtools::session_info() for better results
writeLines(
  capture.output(sessionInfo()),
  here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt"))
)

# save package citations
knitr::write_bib(c(packages, "base"), file = here("software.bib"))

# set up default knitr parameters
# https://yihui.org/knitr/options/
knitr::opts_chunk$set(
  echo = FALSE, # Show outputs, but not code. Change to TRUE to show code as well
  fig.retina = 4,
  fig.width = 8,
  fig.path = paste0(here("results", "figures"), "/")
)
```

## Data and variable

### American Community Survey (ACS) five-year estimate (2014-2018)

- `Title`: American Community Survey (ACS) five-year estimate (2014-2018)
- `Abstract`: Sociodemographic breakdown of disabled population
- `Spatial Coverage`: United States
- `Spatial Resolution`: County
- `Spatial Representation Type`: Vector MULTIPOLYGON
- `Spatial Reference System`: EPSG 4269
- `Temporal Coverage`: 2014-2018
- `Temporal Resolution`: five-year estimate
- `Lineage`: Pulled documented variables from S1810 and C18130 tables using tidyCensus
- `Distribution`: Publicly Available
- `Constraints`: Public Data

The American Community Survey (ACS) five-year estimate (2014-2018) variables used in the study are outlined in the table below.
Details on ACS data collection can be found at <https://www.census.gov/topics/health/disability/guidance/data-collection-acs.html> and details on sampling methods and accuracy can be found at <https://www.census.gov/programs-surveys/acs/technical-documentation/code-lists.html>.

|                            Variable Name in Study                            |                                               ACS Variable name                                                |
|:----------------------------------:|:----------------------------------:|
| percent of total civilian non-institutionalized population with a disability |                                                 S1810_C03_001E                                                 |
|                                   **Race**                                   |                                                                                                                |
|                      percent w disability: White alone                       |                                                 S1810_C03_004E                                                 |
|                      percent w disability: Black alone                       |                                                 S1810_C03_005E                                                 |
|                    percent w disability: Native American                     |                                                 S1810_C03_006E                                                 |
|                      percent w disability: Asian alone                       |                                                 S1810_C03_007E                                                 |
|                       percent w disability: Other race                       |                                                 S1810_C03_009E                                                 |
|                                **Ethnicity**                                 |                                                                                                                |
|                   percent w disability: Non-Hispanic White                   |                                                S1810_C03_0011E                                                 |
|                        percent w disability: Hispanic                        |                                                 S1810_C03_012E                                                 |
|                 percent w disability: Non-Hispanic non-White                 | (S1810_C02_001E - S1810_C02_011E - S1810_C02_012E) / (S1810_C01_001E - S1810_C01_011E - S1810_C01_012E) \* 100 |
|                       percent w disability: Other race                       |                                                 S1810_C03_009E                                                 |
|                                 **Poverty**                                  |                                                                                                                |
|                  percent w disability: Below poverty level                   |                         (C18130_004E + C18130_011E + C18130_018E) / C18130_001E \* 100                         |
|                  percent w disability: Above poverty level                   |                         (C18130_005E + C18130_012E + C18130_019E) / C18130_001E \* 100                         |
|                                   **Age**                                    |                                                                                                                |
|                          percent w disability: 5-17                          |                                                 S1810_C03_014E                                                 |
|                         percent w disability: 18-34                          |                                                 S1810_C03_015E                                                 |
|                         percent w disability: 35-64                          |                                                 S1810_C03_016E                                                 |
|                         percent w disability: 65-74                          |                                                 S1810_C03_017E                                                 |
|                          percent w disability: 75+                           |                                                 S1810_C03_018E                                                 |
|                              **Biological sex**                              |                                                                                                                |
|                          percent w disability: male                          |                                                 S1810_C03_001E                                                 |
|                         percent w disability: female                         |                                                 S1810_C03_003E                                                 |

: Disability Subgroup Variables

American Community Survey (ACS) data for sociodemographic subcategories of people with disabilities can be accessed by using the `tidycensus` package to query the Census API. This requires an API key which can be acquired at [api.census.gov/data/key_signup.html](https://api.census.gov/data/key_signup.html).

```{r}
acs_subject_vars <- load_variables(2018, "acs5/subject")
acs_vars_S1810 <- acs_subject_vars |> filter(str_detect(name, "S1810"))
write_csv(acs_vars_S1810, here("data", "metadata", "acs_S1810_vars.csv"))

acs_vars <- load_variables(2018, "acs5")
acs_vars_C18130 <- acs_vars |> filter(str_detect(name, "C18130"))
write_csv(acs_vars_C18130, here("data", "metadata", "acs_C18130_vars.csv"))
```


```{r API-Load-ACS, eval=FALSE}
# If you wish to use a census API key, run the census_api_key() function in the console

# Query disability demographic data with geographic boundaries
acs <- get_acs(
  geography = "county",
  table = "S1810",
  year = 2018,
  output = "wide",
  cache_table = TRUE,
  geometry = TRUE,
  keep_geo_vars = TRUE
)

# Query poverty and disability data
acs_pov <- get_acs(
  geography = "county",
  table = "C18130",
  year = 2018,
  output = "wide",
  cache_table = TRUE
)

# Query state geographic data
state <- get_acs(
  geography = "state",
  year = 2018,
  variables = c("B01001_001"),
  geometry = TRUE,
  keep_geo_vars = TRUE
)

# Save query results
saveRDS(acs, here("data", "raw", "public", "acs.RDS"))
saveRDS(acs_pov, here("data", "raw", "public", "acs_pov.RDS"))
saveRDS(state, here("data", "raw", "public", "state.RDS"))
```

```{r load-acs}
acs <- readRDS(here("data", "raw", "public", "acs.RDS"))
acs_pov <- readRDS(here("data", "raw", "public", "acs_pov.RDS"))
state <- readRDS(here("data", "raw", "public", "state.RDS"))
```

## County Level COVID-19 Incidence Rate

- `Title`: County Level COVID-19 Incidence Rate
- `Abstract`: Socioodemographic breakdown of disabled population
- `Spatial Coverage`: United States
- `Spatial Resolution`: County
- `Spatial Representation Type`: Vector MULTIPOLYGON
- `Spatial Reference System`: EPSG 4269
- `Temporal Coverage`: 2020-01-22 — 2020-08-01
- `Temporal Resolution`: 8 months
- `Lineage`: Center for System Science in Engineering at Johns Hopkins University August, 01, 2020
- `Distribution`: see below
- `Constraints`: Public Data

Data on COVID-19 cases from the Johns Hopkins University dashboard have been provided directly with the research compendium because the data is no longer available online in the state in which it was downloaded on August 1, 2020.
The dashboard and cumulative counts of COVID-19 cases and deaths were continually updated, so an exact reproduction required communication with the original author, Jayajit Chakraborty, for assistance with provision of data from August 1, 2020.
The data includes an estimate of the total population (`POP_ESTIMA`) and confirmed COVID-19 cases (`Confirmed`).
The COVID-19 case data expresses cumulative count of reported COVID-19 from 1/22/2020 to 8/1/2020.
Although metadata for this particular resource is no longer available from the original source, one can reasonably assume that the total population estimate was based on the 2014-2018 5-year ACS estimate, as the 2019 estimates data had not been released yet.

Versions of the data can be found at the John Hopkins CCSE COVID-19 Data Repository (<https://github.com/CSSEGISandData/COVID-19>).
However, archived data only provides summaries at the national scale.
We received the COVID-19 case data through 8/1/2020 at the county level from the author, as there is no readily apparent way to access archived data from the Johns Hopkins University Center for Systems Science Engineering database.

```{r load-covid-data}
covid <- read_sf(here("data", "raw", "public", "covidcase080120.gpkg"))

# select and rename the fips code, population, cases, and x,y coordinates
covid <- select(covid,
  fips = FIPS,
  pop = POP_ESTIMA,
  cases = Confirmed,
  x = X, y = Y
)
```

## Data transformations

![Workflow](`r here("procedure", "code", "spatialClusteringWorkflow.pdf")`)

### ACS data transformations

The original study extent is the lower 48 states and Washington D.C. Therefore, Alaska, Hawai'i and Puerto Rico are removed from the data (workflow step 1).
Data on people with disabilities in poverty is derived from a different census table (C18130) than data on people with disabilities and age, race, ethnicity, age, and biological sex (S1810).
Therefore, join the poverty data to the other data using the GEOID (workflow step 3).
Also transform the ACS geographic data into Contiguous USA Albers Equal Area projection and fix geometry errors.

```{r filter-join-acs}
# Remove Alaska, Hawaii & Puerto Rico,
# transform coordinate system and fix geometries
acs <- filter(acs, !STATEFP %in% c("02", "15", "72")) %>%
  st_transform(5070) %>%
  st_make_valid()

# Remove Alaska, Hawaii & Puerto Rico,
state <- filter(state, !STATEFP %in% c("02", "15", "72")) %>%
  st_transform(5070)

# Join poverty data to disability data
acs <- left_join(acs, acs_pov, by = "GEOID")
rm(acs_pov)
```

Optionally, save the raw ACS data to `data/raw/public/acs.gpkg` for use in GIS software.

```{r save-ACS, message = F, eval = FALSE}
# Save downloaded acs data to acs.gpkg
st_write(
  acs,
  here("data", "derived", "public", "acs.gpkg"),
  layer = "acs"
)
st_write(
  state,
  here("data", "derived", "public", "acs.gpkg"),
  layer = "state"
)
```

Calculate independent socio-demographic variables of people with disabilities as percentages for each sub-category of disability (race, ethnicity, poverty, age, and biological sex) and remove raw census data from the data frame (workflow step 4).
Reproject the data into an Albers equal area conic projection.

```{r Preprocess-ACS}
# calculate percentages
acs_derived <- mutate(acs,
  dis_pct = S1810_C02_001E / S1810_C01_001E * 100,
  white_pct = S1810_C02_004E / S1810_C01_001E * 100,
  black_pct = S1810_C02_005E / S1810_C01_001E * 100,
  native_pct = S1810_C02_006E / S1810_C01_001E * 100,
  asian_pct = S1810_C02_007E / S1810_C01_001E * 100,
  other_pct =
    (S1810_C02_008E + S1810_C02_009E + S1810_C02_010E) / S1810_C01_001E * 100,
  non_hisp_white_pct = S1810_C02_011E / S1810_C01_001E * 100,
  hisp_pct = S1810_C02_012E / S1810_C01_001E * 100,
  non_hisp_non_white_pct =
    (S1810_C02_001E - S1810_C02_012E - S1810_C02_011E) / S1810_C01_001E * 100,
  bpov_pct = (C18130_004E + C18130_011E + C18130_018E) / C18130_001E * 100,
  apov_pct = (C18130_005E + C18130_012E + C18130_019E) / C18130_001E * 100,
  pct_5_17 = S1810_C02_014E / S1810_C01_001E * 100,
  pct_18_34 = S1810_C02_015E / S1810_C01_001E * 100,
  pct_35_64 = S1810_C02_016E / S1810_C01_001E * 100,
  pct_65_74 = S1810_C02_017E / S1810_C01_001E * 100,
  pct_75 = S1810_C02_018E / S1810_C01_001E * 100,
  male_pct = S1810_C02_002E / S1810_C01_001E * 100,
  female_pct = S1810_C02_003E / S1810_C01_001E * 100
)

# select only relevant geographic identifiers and derived percentages
acs_derived <- acs_derived %>%
  select(
    fips = GEOID,
    statefp = STATEFP,
    county = NAME.x,
    county_st = NAME,
    contains("pct")
  )
```

### COVID-19 data transformations

Calculate the COVID incidence rate as the cases per 100,000 people (workflow step 2).
Convert the COVID data to a non-geographic data frame.

```{r covid-rate}
covid_table <- covid %>%
  mutate(covid_rate = round(covid$cases / covid$pop * 100000, 2)) %>%
  st_drop_geometry()
```

Join dependent COVID data to independent ACS demographic data.

```{r join-covid-to-acs}
# Join COVID incidence rate data to acs data
acs_covid <- acs_derived %>%
  left_join(covid_table, by = "fips")

# move covid_rate column prior to disability percentages
acs_covid <- acs_covid %>%
  select(fips, statefp, county, county_st, covid_rate, everything())

rm(acs, acs_derived, covid)
```

## Missing data

**Unplanned deviation for reproduction**: There is one county with missing disability and poverty data.
This was not mentioned in the original study or in our pre-analyis plan.
However, we replace the missing data with zeros, producing results identical to Chakraborty's.

```{r missing data}
# county with missing data
filter(acs_covid, is.na(bpov_pct)) %>% st_drop_geometry() %>% kable()

# replace NA with 0 for missing data
acs_covid[is.na(acs_covid$bpov_pct), ]$bpov_pct <- 0
acs_covid[is.na(acs_covid$apov_pct), ]$apov_pct <- 0
```

## Map COVID-19 incidence

Map the county level distribution of COVID-19 incidence rates, comparing to Figure 1 of the original study.

```{r map-covid-rates, message = FALSE}
tm_covid_rates <- tm_shape(acs_covid) +
  tm_polygons("covid_rate",
    title = "COVID-19 Cases per 100,000 people\n(22 January 2020 to 1 August 2020)",
    style = "quantile",
    border.alpha = .2,
    lwd = 0.2,
    palette = "Oranges",
  ) +
  tm_shape(state) +
    tm_borders("grey", lwd = .5) +
  tm_layout(
    legend.position = c("left", "bottom"),
    legend.title.size = 0.8,
    legend.text.size = 0.5
  )

tm_covid_rates
```

## Map disability rates

**Unplanned deviation for reproduction**: We also map the spatial distribution of the percent of people with any disability to improve our understanding of the geographic patterns and relationships of between the overarching independent variable (percentage of people with disability) and the dependent variable (COVID-19 incidence rate).

```{r map-disability-rates, message = FALSE}
tm_disability_rates <- tm_shape(acs_covid) +
  tm_polygons("dis_pct",
    title = "Percent of People with Disability\n(ACS 2014-2018)",
    style = "quantile",
    border.alpha = .2,
    lwd = 0.2,
    palette = "Oranges"
  ) +
  tm_shape(state) +
  tm_borders("grey", lwd = .5) +
  tm_layout(
    legend.position = c("left", "bottom"),
    legend.title.size = 0.8,
    legend.text.size = 0.5
  )

tm_disability_rates
```

## Descriptive statistics

Calculate descriptive statistics for dependent COVID-19 rate and independent socio-demographic characteristics, reproducing the min, max, mean, and SD columns of original study table 1.

**Planned deviation for reanalysis**: We also calculate the Shapiro Wilk test for normality.

```{r descriptive-statistics}
acs_covid_stats <- acs_covid %>%
  st_drop_geometry() %>%
  select(covid_rate, contains("pct")) %>%
  stat.desc(norm = TRUE) %>%
  round(2) %>%
  t() %>%
  as.data.frame() %>%
  select(min, max, mean, SD = std.dev, ShapiroWilk = normtest.W, p = normtest.p)

acs_covid_stats %>%
  kable(caption = "Reproduced Descriptive Statistics",
        align = "c") %>%
  column_spec(2:6, width_min = "5em") %>%
  column_spec(7, width_min = "2em") %>%
  kable_styling(full_width = FALSE)
```

Compare reproduced descriptive statistics to original descriptive statistics.
Difference is calculated as 'reproduction study - original study'.
Identical results will result in zero.

```{r compare-descriptive-stats}
# load original table 1 results
table1 <- read.csv(here("data", "raw", "public", "chakraborty", "table1.csv"))

# subtract original results from reproduced results
(select(acs_covid_stats, min, max, mean, SD) -
  select(table1, min, max, mean, SD)) %>%
  kable(caption = "Descriptive Statistics Comparison",
        align = "c") %>%
  column_spec(2:5, width = "4em") %>%
  kable_styling(full_width = FALSE)

rm(acs_covid_stats)
```

The descriptive statistics are identical, except that the original study seems to have rounded the COVID-19 statistics to zero decimal places.

## Analysis

![Workflow](`r here("procedure", "code", "spatialClusteringWorkflow.pdf")`)

# Results

Describe how results are to be presented.

# Discussion

Describe how the results are to be interpreted *vis a vis* each hypothesis or research question.

# Integrity Statement

Include an integrity statement - The authors of this preregistration state that they completed this preregistration to the best of their knowledge and that no other preregistration exists pertaining to the same hypotheses and research.
If a prior registration *does* exist, explain the rationale for revising the registration here.

# Acknowledgements

- `Funding Name`: name of funding for the project
- `Funding Title`: title of project grant
- `Award info URI`: web address for award information
- `Award number`: award number

This report is based upon the template for Reproducible and Replicable Research in Human-Environment and Geographical Sciences, DOI:[10.17605/OSF.IO/W29MQ](https://doi.org/10.17605/OSF.IO/W29MQ)

# References
